Debugging by \textit{lastChange}

### 2010-12-08 papers/ecoop2011/ecoop2011-lastChange.tex

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{subfigure} %[tight,normalsize]
\usepackage{algorithmic}
\usepackage{algorithm}

@@

By browsing through the code or other means\cite{Barton2010},
 the developer determines that the value displayed on the 
button is set at line 22. Since the displayed value is incorrect we know 
the bug occured before we hit this line.
To start debugging, the developer sets a breakpoint
on line 22. Once the button is clicked, execution is paused at line
22. Figure~\ref{fig:example1} shows the Firebug debugger while the
execution is paused. Firebug has several panels (e.g., HTML, CSS,
Script, DOM, etc.) that each demonstrate one aspect of the Web page.
The Script panel contains the list of all loaded source
files and regular debugging facilities such as setting breakpoints and
stepping. To the right of the script panel, the Watch panel shows the program state
where the developer can examine object and variable values. In our case, the
\texttt{myObject.myProperty} value at the paused point is zero. We expected this value to be \texttt{2}.

@@

\begin{figure}[htp]
\centering 
\includegraphics[height=.2\textheight]{6-foo-changes1.jpg} 
\caption{A hypothetical list of  change events for a property \texttt{foo}. 
Each change event adds a column with the id of the object changed, the call stack,
and some program state such as local variable values. 
At the reproduction point  we determine which id  corresponds to 
object \texttt{bar} and read out column 4,  the last
  change of \texttt{bar.foo}. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but its not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

@@

\begin{figure}[htp]
\centering 
\includegraphics[height=.12\textheight]{7-foo-changes2.jpg}
\caption{A hypothetical list of variable \texttt{foo} change events. Each column of the list indicates a change event; for each
change event the scope id returns by \texttt{ScopeId()} is recorded along with call stack and program state information. 
   The last column having the scope id \texttt{foo} at the reproduction point indicates 
  the last change.}
\label{fig:foo-changes2}
\end{figure}

@@

\FOR{$q$ in $lastChange queries$} 
 \FOR {$p$ is defined at the $q$ result}
   \IF {$p$ is a lastChange on object property} 
     \STATE the property owner id should be stored at $q$ change events. 
   \ELSIF {$p$ is a lastChange on variable} 
     \STATE the variable scope id should be stored at $q$ change events.
	 \ENDIF 
 \ENDFOR 
\ENDFOR

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented it in an extension to the Firebug
JavaScript debugger\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute. However the JavaScript prototype is
easy to explore, change, and share with others for feedback.
  A professionally useful debugger would implement these primitive operations within the JavaScript engine.

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. This use of a regular object property  might change
the program behaviour. For example it will apear as an extra property in 
\texttt{for(property in object)} loops. 
The next version of Firefox\footnote[1]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}.
provides an alternative via \texttt{defineProperty} a standard JavaScript
function which receives 
an argument specifying whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object and its predefined properties is available.  For 
our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[2]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox 
JavaScript engine, we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and parse the source code to determine which object was created.

@@

\section{User Study}
We supplied four experienced Javascript developers with our prototype lastChange in an 
extended Firebug debugger. Following a tutorial and a practice case, we observed as they 
applied both conventional breakpoint and \textit{lastChange} on two small programs we 
provided. (Our prototype did not have a user interface to support Sec.~\ref{sec:pausing}).
All four developers successfully applied \textit{lastChange} to the test programs 
and understood how it could help debugging.

We designed the example programs to require two steps of \textit{lastChange}; 
all four users took more steps (5, 10, 10, and more than 25) to find the defect location with breakpoints.  
To find locations for breakpoints two users scrolled through the source, 
another searched the text, a third set a lot of breakpoints to get an understand of the control flow. 
Based on our own experience we expect these strategies represent the kinds of approaches 
developers have available. These operations are time consuming and tedious: all four users 
found the defect location with just two \textit{lastChange} operations. Based on our measurements,
this makes \textit{lastChange} about three times faster than breakpoints for the navigation from 
the reproduction point to the defect.
Of course these programs were designed to highlight lastChange and many kinds of debugging 
issues have been hidden by the design of our tests. Our results show that, when a defect 
relates to incorrect values and a developer recognizes this, then the operational 
mechanics of lastChange lead to the defect much more quickly than breakpoints.

@@

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their mind, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

@@

We have described our prototype all-JavaScript implementation in Sec. 4. It is adequate for 
exploring the ideas and may even be usable in production. However significant 
improvments can be made. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChangeHook()}. Many
object-oriented runtimes provide object identifiers[] and some
provide access to object creation[]. In the particular case of
the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers []. In the case of
tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
JavaScript objects-support \texttt{watch()} function.

@@

\begin{thebibliography}{16}
\bibitem{Bond}%[Bond(2007)]
M.D. Bond, N. Nethercote, S.W. Kent, S.Z. Guyer, and K.S. McKinley. \newblock Tracking bad apples: reporting the origin of null and undefined value errors.
\newblock In \emph{22nd annual ACM SIGPLAN conference on Object-oriented programming, systems, languages, and applications(OOPSLA)},
October, 2007.

@@

\bibitem{Firebug}%[Firebug(2010)]
Firebug. \newblock http://getfirebug.com.

\bibitem{Firefox}%[Firefox(2010)]
Firefox. \newblock http://www.mozilla.com.

@@

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{subfigure} %[tight,normalsize]
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{textcomp}

@@

By browsing through the code or other means\cite{Barton},
 the developer determines that the value displayed on the 
button is set at line 22. Since the displayed value is incorrect we know 
the bug occured before we hit this line.
To start debugging, the developer sets a breakpoint
on line 22. Once the button is clicked, execution is paused at line
22. Figure~\ref{fig:example1} shows the Firebug debugger while the
execution is paused. Firebug has several panels (e.g., HTML, CSS,
Script, DOM, etc.) that each demonstrate one aspect of the Web page.
The Script panel contains the list of all loaded source
files and regular debugging facilities such as setting breakpoints and
stepping. To the right of the script panel, the Watch panel shows the program state
where the developer can examine object and variable values. In our case, the
\texttt{myObject.myProperty} value at the paused point is zero. We expected this value to be \texttt{2}.

@@

\begin{figure}[htp]
\centering 
\includegraphics[height=.23\textheight]{6-foo-changes1.jpg} 
\caption{A hypothetical list of  change events for a property \texttt{foo}. 
Each change event adds a column with the id of the object changed, the call stack,
and some program state such as local variable values. 
At the reproduction point  we determine which id  corresponds to 
object \texttt{bar} and read out column 4,  the last
  change of \texttt{bar.foo}. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but its not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

@@

\begin{figure}[htp]
\centering 
\includegraphics[height=.14\textheight]{7-foo-changes2.jpg}
\caption{A hypothetical list of variable \texttt{foo} change events. Each column of the list indicates a change event; for each
change event the scope id returns by \texttt{ScopeId()} is recorded along with call stack and program state information. 
   The last column having the scope id \texttt{foo} at the reproduction point indicates 
  the last change.}
\label{fig:foo-changes2}
\end{figure}

@@

\FOR{$q$ in $lastChange$ queries} 
 \FOR {$p$ is defined at the $q$ result}
   \IF {$p$ is a lastChange on object property} 
     \STATE the property owner id should be stored at $q$ change events. 
   \ELSIF {$p$ is a lastChange on variable} 
     \STATE the variable scope id should be stored at $q$ change events.
	 \ENDIF 
 \ENDFOR 
\ENDFOR

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented\footnote[1]{http://code.google.com/p/querypoint-debugging} it in an extension to the Firebug
JavaScript debugger\footnote[2]{http://getfirebug.com}. %\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser.%\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute. However the JavaScript prototype is
easy to explore, change, and share with others for feedback.
  A professionally useful debugger would implement these primitive operations within the JavaScript engine.

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. This use of a regular object property  might change
the program behaviour. For example it will apear as an extra property in 
\texttt{for(property in object)} loops. 
The next version of Firefox\footnote[3]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}.
provides an alternative via \texttt{defineProperty} a standard JavaScript
function which receives 
an argument specifying whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object and its predefined properties is available.  For 
our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[4]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox 
JavaScript engine, we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and parse the source code to determine which object was created.

@@

\section{User Study}
We supplied four experienced Javascript developers with our prototype lastChange in an 
extended Firebug debugger\footnote[5]{http://ltiwww.epfl.ch/\texttildelow mirghase/lastchange-userstudy}. Following a tutorial and a practice case, we observed as they 
applied both conventional breakpoint and \textit{lastChange} on two small programs we 
provided. (Our prototype did not have a user interface to support Sec.~\ref{sec:pausing}).
All four developers successfully applied \textit{lastChange} to the test programs 
and understood how it could help debugging.

We designed the example programs to require two steps of \textit{lastChange}; 
all four users took more steps (9, 13, 12, and more than 30) to find the defect location with breakpoints.  
To find locations for breakpoints two users scrolled through the source, 
another searched the text, a third set a lot of breakpoints to get an understand of the control flow. 
Based on our own experience we expect these strategies represent the kinds of approaches 
developers have available. These operations are time consuming and tedious: all four users 
found the defect location with just two \textit{lastChange} operations. Based on our measurements,
this makes \textit{lastChange} about three times faster than breakpoints for the navigation from 
the reproduction point to the defect.
Of course these programs were designed to highlight lastChange and many kinds of debugging 
issues have been hidden by the design of our tests. Our results show that, when a defect 
relates to incorrect values and a developer recognizes this, then the operational 
mechanics of lastChange lead to the defect much more quickly than breakpoints.

@@

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change[].  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their mind, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

@@

We have described our prototype all-JavaScript implementation in Sec. 4. It is adequate for 
exploring the ideas and may even be usable in production. However significant 
improvments can be made. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChan geHook()}. Many
object-oriented runtimes provide object identifiers and 
provide access to object creation directly or by bytecode instrumentation\cite{JPDA}. 
In the particular case of the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers []. In the case of
tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
JavaScript objects-support \texttt{watch()} function.

@@

\begin{thebibliography}{16}

\bibitem{Barton}%[Bond(2010)]
J.J. Barton, and J. Odvarko. \newblock Dynamic and graphical web page breakpoints.
\newblock In \emph{Conference on World Wide Web(WWW)},
April, 2010.

\bibitem{Bond}%[Bond(2007)]
M.D. Bond, N. Nethercote, S.W. Kent, S.Z. Guyer, and K.S. McKinley. \newblock Tracking bad apples: reporting the origin of null and undefined value errors.
\newblock In \emph{22nd annual ACM SIGPLAN conference on Object-oriented programming, systems, languages, and applications(OOPSLA)},
October, 2007.

@@

%\bibitem{Firebug}%[Firebug(2010)]
%Firebug. \newblock http://getfirebug.com.

%\bibitem{Firefox}%[Firefox(2010)]
%Firefox. \newblock http://www.mozilla.com.

@@

\bibitem{JPDA}%[JSD(2010)]
Java Platform Debugger Architecture. \newblock http://java.sun.com/javase/technologies/ core/toolsapis/jpda.
